The role of cognateness in native spoken word recognition

### 2024-07-11 manuscript/manuscript.tex
Fix PDF output

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{Times}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in non-native spoken word recognition},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and non-native word
recognition }
\title{The role of cognateness in non-native spoken word recognition}
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}\And\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}}\And\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}}\And\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Understanding spoken words in a non-native language is easier when the
words are phonologically similar to their translations into the
listener's native language are phonologically similar. Phonological
similarity plays a more important role in the language performance of
low-proficiency bilinguals, compared to that of high-proficiency
bilinguals., and This suggests that low-proficiency bilinguals may
points to the latter as relying more strongly on cross-language
word-to-word links (lexical route), while high-proficiency bilinguals
rely more on word-concept associations (conceptual route) than on
cross-language word-to-word links (lexical route). Disentangling
bilinguals' reliance on the conceptual and lexical and conceptual routes
during translation is challenging, given that both sources of
information are present during spoken word comprehensionto some extent
in both high- and low-proficiency bilinguals. In this study, we tested
English and Spanish native speakers in a translation elicitation task,
in which they listened to words in an unfamiliar language (Spanish or
Catalan for English natives, and Catalan for Spanish natives), and then
had to translate them to their native language. Given their lack of
previous knowledge ion the unfamiliar tested language, phonological
similarity between the presented words and their correct translations
was the only cue available for participants to succeed in the task. This
allowed us to explore the informativeness of the lexical route for word
translation, in the absence of information from the conceptual route.
Our results indicate that participants benefited from the phonological
similarity between translation equivalents: the more similar the word
they listened to and its translational equivalent, the higher the
probability of a correct translation. These results point to
cross-language word-to-word links as providing rich information during
word translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

Listening to non-native speech is costlier than listening to native
speech, even for highly proficient bilinguals, and especially in
acoustically adverse situations
(\citeproc{ref-lecumberri2010non}{Lecumberri et al., 2010};
\citeproc{ref-takata1990english}{Takata \& Nábělek, 1990}). However,
there are some sources of information which may make non-native speech
comprehension easier. One of the sources of this increased difficulty
stems from the possible mismatch between the phonology of the native and
the non-native languages: some acoustic features embedded in the
non-native speech signal do not overlap perfectly with any phonemic
category in the listener's native language. For example, imagine a
Spanish native, with no previous familiarity with any other language,
who listens for the first time to French may encounter the word /pɔʁt/
(\emph{porte}), translation of door in French. The voiced uvular
fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/ do not
exist in Spanish as phonemic categories. These discrepancies can
nonetheless be perceived as allophonic variations the native (Spanish)
phonemes /r/ and /o/ (\citeproc{ref-best2001discrimination}{Best et al.,
2001}), and the non-native speech signal may engage lexical processing
mechanisms (\citeproc{ref-weber2004lexical}{Weber \& Cutler, 2004}),
leading to the activation and selection, and therefore the recognition
of its translation to Spanish /ˈpwer.ta/ (\emph{puerta}). The
recognition of this French, unfamiliar word relies almost entirely on
its similarity with its Spanish translation. Conversely, imagine now
that the same Spanish listener hears English for the first time and
encounters the word /dɔː/ (\emph{door}). This time, phonological
similarity is of little help: /ˈpwer.ta/ and /dɔː/ share very few
phonemes. It would come as a surprise if the listener was able to
translate the English word successfully, if not by relying on previously
learnt word-concept associations (which they lack, as they are
unfamiliar to English).

Cross-language phonological similarity seems, at face value, an
important source of information for unfamiliar listeners during auditory
word recognition, even if translations are not phonologically identical.
Although this mismatch has a noticeable toll on comprehension
(\citeproc{ref-cutler2004patterns}{Cutler et al., 2004}), the fact that
word recognition can take place at all in such circumstances illustrates
that non-native listeners--even low-proficiency ones--are rarely
completely naïve to the language they are listening to. In the present
study, we investigate the extent to which adult listeners rely on the
phonological similarity between their native and non-native languages
during word recognition, capitalising on the role of \emph{cognates.}

Many languages share, to some extent, similarities at the lexical level.
This is frequently due to their typological closeness and/or
socio-historical events involving the speakers of these languages (e.g.,
migration, social contact). Cognates embody a particular case of such
commonalities. Cognates are defined as cross-language synonyms which
have similar forms whose form is similar (at the phonological level,
orthographic level, signed level), commonly due to their shared
etymological origin\footnote{Some form-similar cross-language synonyms
  are technically not cognates. For example, \emph{sun} and \emph{sol}
  (in Spanish) share phonological onset but their etymology points to
  different origins. The complimentary case is equally problematic when
  defining cognateness: cross-language synonyms that share etymological
  origin may not be necessarily be recognised as cognates by the average
  listener. Such is the case of \emph{cheese} and \emph{queso} which
  share Latin root but are phonologically dissimilar. For simplicity, we
  will use the term \emph{cognateness} in this paper to include all
  cross-language synonyms that fulfill a specified threshold of form
  similarity, regardless of etymology. We do not expect that etymology
  plays a direct role on language perception if it is not via
  form-similarity, since it is not necessary for participants in
  psycholinguistic experiments to be aware of the etymology of the words
  they encounter in the tasks to be subject to the effect of
  form-similarity.}. For example, Romance languages such as Spanish and
Catalan share many cognates
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (door in Spanish and Catalan,
respectively). Cognates play a pivotal role in many models of bilingual
lexical processing because they provide evidence that bilinguals access
the lexicon in a language non-selective way. For instance, pictures are
named faster by Spanish-Catalan bilinguals when their associated labels
in both languages are cognates (e.g., \emph{puerta}-\emph{porta}),
compared to when their labels are non-cognates
(\emph{mesa}-\emph{taula}, {[}table{]}), which suggests that word
representations of both languages are activated during word retrieval
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}). This cognate
advantage is not restricted to production, but also extends to word
recognition (\citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), word learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022cross}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and word
translation (\citeproc{ref-christoffels2006memory}{Christoffels et al.,
2006}).

\subsection{Lexical and conceptual routes for
translation}\label{lexical-and-conceptual-routes-for-translation}

Early theoretical accounts of this `Cognate effect' proposed that the
impact of cognateness on lexical processing depends on bilinguals'
proficiency in their second language (L2)
(\citeproc{ref-potter1984lexical}{Potter et al., 1984}; see
\citeproc{ref-chen1989patterns}{Chen \& Leung, 1989} for similar
results). The reasoning behind this claim was the assumption that
second-language learners start acquiring words in their L2 by
associating them to their translation in their L1 (lexical route),
instead of directly to their shared concept (conceptual route). This
word-to-word connectivity should be sensitive to form-similarity (i.e.,
cognateness): the more similar the two word- forms are, the stronger the
connection. As learners become more proficient, the connection between
L2 representations and their meaning grows stronger, and L2 word
processing becomes less reliant on word-word connections between L1 and
L2 representations. Cognates subsequently exert less impact on lexical
processing in proficient speakers (see
\citeproc{ref-andras2022cognate}{Andras et al., 2022} for recent
experimental evidence for this claim). The Revised Hierarchical Model
(RHM, \citeproc{ref-kroll1994category}{Kroll \& Stewart, 1994}) captured
this hypothesis and predicted that translating words from L1 to L2
should take longer than translating from L2 to L1. The rationale behind
this prediction was that L1-to-L2 translation relies more strongly on
word-to-word links between L1 and L2 representations, where L2-to-L1
translation relies more strongly on the mediation between the concept
and the two word- forms. L1-to-L2 translation should therefore be more
sensitive to the form similarity between the L1 and the L2
representations: cognate words should be retrieved faster than
non-cognate words during L1-to-L2 translation.

To test these predictions, Groot et al.
(\citeproc{ref-degroot1994forward}{1994}) and Groot
(\citeproc{ref-de1992determinants}{1992}) asked Dutch natives with high
(but non-native) English proficiency to translate Dutch words to English
(L1-to-L2) or English words to Dutch (L2-to-L1). Although participants'
performance in both conditions was roughly equivalent for cognates,
non-cognates were translated more slowly from Dutch to English than from
English to Dutch, suggesting that L1-to-L2 translation is more sensitive
to form-similarity than L2-to-L1 translation. Conversely, participants'
performance was more sensitive to semantic variables (e.g.,
concreteness) when translating English words to Dutch, compared to when
translating Dutch words to English, suggesting that L2-to-L1 translation
was more sensitive to the shared conceptual properties of the
translation pair.

These findings support a soft version of the RHM model, in which both
the conceptual and the lexical translation routes are active during
translation, but where L1-to-L2 translation relies more strongly on
word-to-word links between L1 and L2 representations than L2-to-L1
translation does, and therefore is more sensitive to cognateness.
Further evidence for the presence of this difference between L1-to-L2
and L2-to-L1 translations on the degree of reliance on the lexical and
or the conceptual routes between L1-to-L2 and L2-to-L1 translation was
provided by Phillips et al. (\citeproc{ref-phillips2006erp}{2006}), who
used eletrophysiological recordings of bilinguals translating words from
L1 to L2 or \emph{vice versa} after semantic or phonological priming.
The authors reported stronger N400 responses (associated with semantic
priming) in L1-to-L2 translation than in L2-to-L12 translation, and
stronger phonological mismatch negativity responses (associated with
phonological priming) during L2-to-L1 translation, compared to L1-to-L2
translation.

\subsection{Competition from lexical
neighbours}\label{competition-from-lexical-neighbours}

Since the RHM model was proposed, later studies on monolingual
populations have focused how on the role of the network of connections
that words establish networks of connections with each other at the
form-level (phonological or orthographic) or at the conceptual level.
For instance, the Neighbourhood Activation Model (NAM,
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}) highlighted
how lexical selection is affected by the number of phonological
neighbours that the selected word has. Words surrounded by a larger
number of phonological neighbours (words that only differ in one phoneme
from the target word) are responded to more slowly and less accurately
than words from sparser phonological neighbourhoods
(\citeproc{ref-goldinger1989priming}{Goldinger et al., 1989};
\citeproc{ref-luce1990similarity}{Luce et al., 1990}), especially when
these neighbours' lexical frequency is higher
(\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).

The existence of phonological links between words is supported with
experimental evidence from the visual world paradigm and phonological
priming. Meyer \& Damian (\citeproc{ref-meyer2007activation}{2007})
showed that target pictures were named faster when presented together
with a phonologically-related distractor, suggesting that the activation
of shared phonological information allowed the target to be activated
and produced faster. there are also findings suggesting that
facilitatory or inhibitory effects on target recognition depend on the
degree of similarity the competitor shares with the target
(\citeproc{ref-dufour2003lexical}{Dufour \& Peereman, 2003};
\citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}). When there is only mismatch of one phoneme between prime and
target, inhibitory effects are exerted on target recognition
(\citeproc{ref-dufour2003lexical}{Dufour \& Peereman, 2003}). This
inhibition was proposed to be due to strong lexical activation of the
prime which results in lexical interference, as opposed to simple
phonological activation from less similar primes that triggers
facilitation.

These findings of phonological links between words in the monolingual
lexicon pose an important question in the field of bilingualism
research: do word representations in one language form part of the
phonological or orthographic neighbours of the other language? Weber \&
Cutler (\citeproc{ref-weber2004lexical}{2004}) observed that bilinguals
displayed lexical competition effects from words in their native
language when performing a word-referent matching task in their L2. This
highlights that lexical competition for bilinguals during language
comprehension may not only come from the tested language but also the
native language, despite the native language not being actively required
for the task. Other studies have similarly found interference when
spoken target words in one language had phonological overlap with a word
in the participant's other language, even when the other language was
not required by the task (\citeproc{ref-desroches2022dynamics}{Desroches
et al., 2022}; \citeproc{ref-spivey1999cross}{Spivey \& Marian, 1999}).
Such patterns have even been found with young children less than 4 years
old (\citeproc{ref-von2012language}{Von Holzen \& Mani, 2012}).

The Bilingual Interactive Activation (BIA/BIA+)
(\citeproc{ref-dijkstra2002architecture}{Dijkstra \& Van Heuven, 2002};
\citeproc{ref-van1998orthographic}{Van Heuven et al., 1998}) addressed
this issue and suggested that lexical representations from both
languages establish both excitatory and inhibitory connections with each
other, resulting in an integrated lexicon for both languages. This is,
in line with connectionist approaches to lexical processing
(\citeproc{ref-mcclelland1981interactive}{McClelland \& Rumelhart,
1981}). Evidence supporting such an integrated lexicon is provided by
the fact that bilinguals' performance in word recognition tasks is
sensitive to the orthographic neighbourhood density of the presented
word's translation. When bilingual participants are presented with words
in one language in a lexical decision task, translating them to the
other language took longer when they were part of large neighbourhoods
(\citeproc{ref-van1998orthographic}{Van Heuven et al., 1998}). This
suggests that presented words activated orthographic neighbourhoods in
the other language, which competed for selection with the target word
during recognition.

\subsection{The present study}\label{the-present-study}

In the present study, we tested listeners' reliance on phonological
similarity when translating words from an unfamiliar language to their
native language, in order to investigate the plausibility of
word-to-word connections during L1-to-L2 translation. We tested English
native speakers and Spanish native speakers in a translation elicitation
task. This paradigm has been used by bilingualism researchers to
identify easily-recognisable cognates for experimental studies. Using
this paradigm, we aim to delve deeper into the factors affecting foreign
word recognition by monolingual participants with no prior exposure to
the language.

Participants were presented with spoken words in an unfamiliar
(non-native language) and were asked to translate them to their native
language. We will henceforth refer to the auditory-presented words heard
by participants on each trial as presented words, and the correct
translation for the presented words as target translations. The
motivation behind testing participants in an unfamiliar language was
that these participants can be considered a particular (extreme) case of
unbalanced bilingualism, in which they do not know any words in L2, and
therefore word-to-concept connections are, in principle, absent. As a
consequence, these participants should only be able use word-to-word
connections to succeed in the translation of words in an unfamiliar
language. We predicted that, if the information provided by the
phonological form of the unfamiliar presented words is sufficient for
translation, participants' performance should increase when the
translation pairs are phonologically more similar.

Following Luce \& Pisoni (\citeproc{ref-luce1998recognizing}{1998})`s
NAM model, we further hypothesised that if participants are able to
correctly translate words from an unfamiliar language based on their
phonological similarity to the target word, similar sounding but
incorrect translations (e.g., false friends) should be activated too. If
this is the case, non-native words that have more phonological
neighbours should be less likely to be correctly translated, especially
if such neighbours have higher lexical frequency and are phonologically
closer to the presented word than the target translation. For instance,
one could expect English participants to incorrectly translate the
Spanish word \emph{botón} as bottom instead of as its correct
translation \emph{button} (because the lexical frequency of
\emph{bottom} is higher than that of \emph{button} and also phonological
closer). This combined constraint led us to calculate a specific form of
phonological neighbours where a neighbour is counted only if it is
higher frequency and has stronger phonological similarity than the
target translation as judged by Levenshtein distance. If the
phonological neighbourhood density of the target translation affects
participants' performance negatively, this would suggest that non-native
words trigger competition between words in the native language during
translation.

In Experiment 1, we collected data from two groups of British English
natives. One group was presented with Catalan words, the other with
Spanish words. We examined the extent to which participants were able to
use the phonological similarity between the presented (Catalan or
Spanish) word and its translation (i.e., cognateness) to provide
accurate responses. In Experiment 2, we tested a group of (European)
Spanish natives in the same task, who were presented with Catalan words.
Catalan and Spanish are both Romance languages whose close typological
distance is reflected in the fact that they share many cognates, where
English is a Germanic language that shares considerably fewer cognates
with Catalan and Spanish. By testing participants translating words from
typologically close and distant languages, we expected to widen the
range of the phonological similarity scores of the translation pairs
involved in the experimental task, therefore allowing us to explore
potential cross-language differences in participant's performance.
Participants in Experiments 1 and 2 were surprisingly good at
translation some words, which had low phonological similarity with their
correct translation. In Experiment 3, we collected additional data from
a new group of British English natives, who were presented with Catalan
and Spanish words. In contrast with the procedure in Experiment 1,
participants in Experiment 3 where asked, after providing a response in
each trial, to report whether they had previous knowledge of the meaning
of the presented word (e.g., some Spanish words are common knowledge,
even for monolingual speakers of English). Finally, we present analyses
on the joint data sets of Experiments 1 to 3.

@@

\captionsetup{labelsep=none}

@@

\captionsetup{labelsep=none}\includegraphics{manuscript_files/figure-pdf/fig-coefs-1.pdf}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{Times}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in native spoken word recognition},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and native word recognition }
\title{The role of cognateness in native spoken word recognition}
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}\And\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}}\And\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}}\And\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Understanding spoken words in a non-native language is easier when their
translations into the listener's native language are phonologically
similar. Phonological similarity plays a more important role in the
language performance of low-proficiency bilinguals, compared to that of
high-proficiency bilinguals. This suggests that low-proficiency
bilinguals may points to the latter as relying more strongly on
cross-language word-to-word links (lexical route), while
high-proficiency bilinguals rely more on word-concept associations
(conceptual route) than on cross-language word-to-word links (lexical
route). Disentangling bilinguals' reliance on the conceptual and lexical
and conceptual routes during translation is challenging, given that both
sources of information are present during spoken word comprehension to
some extent in both high- and low-proficiency bilinguals. In this study,
we tested English and Spanish native speakers in a translation
elicitation task, in which they listened to words in an unfamiliar
language (Spanish or Catalan for English natives, and Catalan for
Spanish natives), and then had to translate them to their native
language. Given their lack of previous knowledge on the unfamiliar
tested language, phonological similarity between the presented words and
their correct translations was the only cue available for participants
to succeed in the task. This allowed us to explore the informativeness
of the lexical route for word translation, in the absence of information
from the conceptual route. Our results indicate that participants
benefited from the phonological similarity between translation
equivalents: the more similar the word they listened to and its
translational equivalent, the higher the probability of a correct
translation. These results point to cross-language word-to-word links as
providing rich information during word translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

Listening to non-native speech is costlier than listening to native
speech, even for highly proficient bilinguals, and especially in
acoustically adverse situations
(\citeproc{ref-lecumberri2010non}{Lecumberri et al., 2010};
\citeproc{ref-takata1990english}{Takata \& Nábělek, 1990}). One of the
sources of this increased difficulty is the partial mismatch between the
phonology of the native and the non-native languages: some acoustic
features embedded in the non-native speech signal do not overlap
perfectly with any phonemic category in the listener's native language.
For example, imagine a Spanish native with no previous familiarity with
any other language listening to French for the first time. This listener
may encounter the word /pɔʁt/ (\emph{porte}), translation of door in
French. The voiced uvular fricative consonant /ʁ/ and the open-mid back
rounded vowel /ɔ/ do not exist in Spanish as phonemic categories. These
discrepancies can nonetheless be perceived as allophonic variations the
native (Spanish) phonemes /ɾ/ and /o/
(\citeproc{ref-best2001discrimination}{Best et al., 2001}), leading to
the non-native speech signal engaging lexical processing mechanisms
(\citeproc{ref-weber2004lexical}{Weber \& Cutler, 2004}). Even if
unfamiliar with the word-form /pɔʁt/, the Spanish monolingual is likely
to activate the phonological representation of its Spanish translation
/pweɾta/ (\emph{puerta}), ultimately accessing the right semantic
representation. Although the phonetic mismatch between the non-native
word and its translation can have a noticeable toll on comprehension
(\citeproc{ref-cutler2004patterns}{Cutler et al., 2004}), the fact that
word recognition can take place at all in such circumstances illustrates
that non-native listeners---even low-proficiency ones---are rarely
completely naïve to the language they are listening to.

Imagine now that the same Spanish listener hears English for the first
time and encounters the word /dɔː/ (\emph{door}). This time,
phonological similarity is of little help: /pweɾta/ and /dɔː/ do not
share any phonemes. It would come as a surprise if the listener was able
to translate the English word successfully. Non-native listeners rely on
the phonological similarity between the word-forms presented in the
input and their translations. Luckily for them, most languages share
such kind of similarity. This is frequently due to their typological
closeness and/or socio-historical events involving the speakers of these
languages (e.g., migration, social contact). For example, Romance
languages such as Spanish and Catalan share many form-similar
translation equivalents (also known as \emph{cognates})
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (door in Spanish and Catalan,
respectively). This type of cross-linguistic similarity has been
extremely informative for the study of bilingual lexical processing:
words that share high form-similarity with their translations are
processed faster and more accurately than words sharing lower
phonological similarity. This phenomenon has provided strong evidence in
favour of the language-non selective hypothesis of bilingual lexical
access, which states that bilinguals activate lexical representations in
both languages, even during monolingual situations. This facilitation
effect has been reported for comprehension
(\citeproc{ref-dijkstra2010cross}{Dijkstra et al., 2010};
\citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}). The
role that cross-linguistic phonological similarity might play in
monolingual lexical access during non-native speech comprehension is
unclear. How much phonological similarity is enough for a monolingual to
understand a word from a non-native language? How does phonological
similarity interact with other determinants of word recognition?

In the present study, we investigate the extent to which monolinguals
can exploit phonological similarity when listening to words in an
unfamiliar languages, in order to achieve comprehension, while taking
into account other lexically-relevant variables interact with the
lexical processing mechanisms involved. We used a translation
elicitation task in which participants listened to words from an
unfamiliar language, and then tried to guess their translation in their
native language. We will henceforth refer to the auditory-presented
words heard by participants on each trial as \emph{presented words}, and
the correct translation for the presented words as \emph{target
translations}. We explored listeners' reliance on phonological
similarity to succeed in the task by manipulating the amount of
phonological similarity between the presented words and their target
translations. Since participants were unfamiliar with the presented
language, they could only be able to use word-to-word connections to
succeed in the translation of words in an unfamiliar language. We
therefore predicted that, if the information provided by the
phonological form of the unfamiliar presented words is sufficient for
translation, participants' performance should increase when the
translation pairs are phonologically more similar.

The extent to which participants can rely on phonological similarity is
limited by other factors. In particular, the presence of phonological
neighbours (word-forms than differ in one phoneme) may interfere with
the activation and selection of the appropriate lexical representation
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
specially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
translation (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).
The inhibitory effect of phonological neighbourhoods has also been
reported across languages. For instance, Weber \& Cutler
(\citeproc{ref-weber2004lexical}{2004}) observed that bilinguals
displayed lexical competition effects from words in their native
language when performing a word-referent matching task in their L2.

In the case of the present task, the potential effect of cross-language
phonological neighbourhood density is critical. The presence of
phonologically similar non-translation equivalents (e.g., false friends)
may act as distractor during lexical access, obstructing the selection
of the appropriate target translation in the listener's lexicon. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with words on
the target language other than its translation, specially if any of them
is more lexically frequent than the target translation. We hypothesised
that the size of the facilitatory role of phonological similarity would
be inversely proportional to the amount of higher-frequency phonological
neighbours of the presented word in the target language. For instance,
one could expect English participants to incorrectly translate the
Spanish word \emph{botón} as bottom instead of as its correct
translation \emph{button} (because the lexical frequency of
\emph{bottom} is higher than that of \emph{button} and also phonological
closer). To test this prediction, we developed a lexical
frequency-dependent measure of cross-language phonological neighbourhood
density, in which a neighbour is counted only if it is higher frequency
and is one phoneme appart from the presented word. If the phonological
neighbourhood density of the target translation affects participants'
performance negatively, this would suggest that non-native words trigger
competition between words in the native language during translation.

In Experiment 1, we collected data from two groups of British English
natives living in the United Kingdom. One group was presented with
Catalan words, the other with Spanish words. We examined the extent to
which participants were able to use the phonological similarity between
the presented word (in Catalan or Spanish) and its target translation to
provide accurate responses. In Experiment 2, we tested a group of
(European) Spanish natives in the same task, who were presented with
Catalan words. Catalan and Spanish are both Romance languages whose
close typological distance is reflected in the fact that they share many
cognates, where English is a Germanic language that shares considerably
fewer cognates with Catalan and Spanish. By testing participants
translating words from typologically close and distant languages, we
expected to widen the range of the phonological similarity scores of the
translation pairs involved in the experimental task, therefore allowing
us to explore potential cross-language differences in participant's
performance. Participants in Experiments 1 and 2 were surprisingly good
at translation some words, which had low phonological similarity with
their correct translation. In Experiment 3, we collected additional data
from a new group of British English natives, who were presented with
Catalan and Spanish words. In contrast with the procedure in Experiment
1, participants in Experiment 3 where asked, after providing a response
in each trial, to report whether they had previous knowledge of the
meaning of the presented word (e.g., some Spanish words are common
knowledge, even for monolingual speakers of English). Finally, we
present analyses on the joint data sets of Experiments 1 to 3.

@@

@@

\includegraphics{manuscript_files/figure-pdf/fig-coefs-1.pdf}
